Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-11 paper/paper.tex
minor style tweaks

The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
\begin{wrapfigure}{r}{.4\linewidth}
  \includegraphics[width=1.\linewidth]{fig_dmn.png}
  \centering
  \caption{\textbf{Morphological coupling between reward system and
  default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11+/-0.02$) and right NAc volume ($R^2=0.14+/-0.02$) could be predicted from volume in the DMN nodes. These out-of-sample generalization performances were obtained from support vector regression applied to normalized node volumes in the DMN in a 10-fold cross-validation procedure. 
  Consistent for the left and right reward system, NAc volume in a given subject is
  positively coupled with the vmPFC and HC.
  Code and data for reproduction and visualization:
  \url{www.github.com/banilo/to_be_added_later}.
  }
  \label{fig:dmn}
\end{wrapfigure}
Such highly abstract neural computations necessarily rely on the
construction of probabilistic internal information drawing from
memory recall, mental scene elaboration processes,
and stored knowledge of world regularities.
%
Moreover,
in a computational neuroimaging experiment,
dorsomedial PFC activity preferentially processed the consequences of
action choices that were considered but not actually executed,
whereas ventromedial PFC (vmPFC) activity
processed especially value outcomes of actions actually performed on the environment
\citep{nicolle2012agent}.

@@

\pagebreak[2]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Neural correlates of the Bellman equation in the DMN:}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN node were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}

@@

On another front, since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}  
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
\end{wrapfigure}
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

>>>>

The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
Such highly abstract neural computations necessarily rely on the
construction of probabilistic internal information drawing from
memory recall, mental scene elaboration processes,
and stored knowledge of world regularities.
\begin{wrapfigure}{r}{.4\linewidth}
  \includegraphics[width=1.\linewidth]{fig_dmn.png}
  \centering
  \caption{\textbf{Morphological coupling between reward system and
  default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11+/-0.02$) and right NAc volume ($R^2=0.14+/-0.02$) could be predicted from volume in the DMN nodes. These out-of-sample generalization performances were obtained from support vector regression applied to normalized node volumes in the DMN in a 10-fold cross-validation procedure. 
  Consistent for the left and right reward system, NAc volume in a given subject is
  positively coupled with the vmPFC and HC.
  Code and data for reproduction and visualization:
  \url{www.github.com/banilo/to_be_added_later}.
  }
  \label{fig:dmn}
\end{wrapfigure}
%
Moreover,
in a computational neuroimaging experiment,
dorsomedial PFC activity preferentially processed the consequences of
action choices that were considered but not actually executed,
whereas ventromedial PFC (vmPFC) activity
processed especially value outcomes of actions actually performed on the environment
\citep{nicolle2012agent}.

@@

\pagebreak[4]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Neural correlates of the Bellman equation in the DMN:}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN node were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}

@@

On another front, since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}  
  \centering
  \includegraphics[width=1.\linewidth]{darkroom_resized.png}
\end{wrapfigure}
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.
